EGGs are adhesive!\\ An algebraic account of equality saturation

### 2025-09-17 jlamp/jlamp.tex
minima

Consider the term rewriting rule such $x / x \to 1$, from the key example in~\cite{WillseyNWFTP21}.
This is a problematic rule to be simulates using classical term graph rewriting, since it is not left-linear
(that is, $x$ occurs twice in the left-hand side)~\cite{xxx}.
Using e-term graphs (variables are represented as nodes that do not occur among the targets of an edge),
 it can be modelled as the DPO rule below, concisely given by the arrow $L \rightarrow R$
\[
	\hspace{.25cm}
	\xymatrix{        
		\begin{tikzpicture}[baseline=(w.base)]\begin{pgfonlayer}{nodelayer}
				\node[style=node](v1) at (-0.9, 0.3){};
				\node[style=node](v2) at (-0.9, -0.3){};
				\node[style=medium box] at (0, 0){$/$};
				\node[style=none](divfst) at (-0.3, 0.3){};
				\node[style=none](divsnd) at (-0.3, -0.3){};
				\node[style=none](divout) at (0.3, 0){};
				\node[style=node] (w) at (0.9, 0){};
			\end{pgfonlayer}        
			\begin{pgfonlayer}{edgelayer}
				\draw(v1) to (divfst.center);
				\draw(v2) to (divsnd.center);
				\draw(divout.center) to (w);
			\end{pgfonlayer}
			\begin{pgfonlayer}{eqlayer}
				\draw[dashed, rounded corners](-1.2, 0.6) rectangle (-0.6, -0.6);
			\end{pgfonlayer}\begin{pgfonlayer}{background}
				\draw[color=white] (-1.5, 0.2) rectangle (1.5, -0.2);
			\end{pgfonlayer}
		\end{tikzpicture}
		\ar@{=>}[r] &
		\begin{tikzpicture}[baseline=(nil.center)]\begin{pgfonlayer}{nodelayer}
				\node[style=node](v1) at (-0.9, 0.3){};
				\node[style=node](v2) at (-0.9, -0.3){};
				\node[style=medium box] at (0, 0){$/$};
				\node[style=none](divfst) at (-0.3, 0.3){};
				\node[style=none](divsnd) at (-0.3, -0.3){};
				\node[style=none](divout) at (0.3, 0){};
				\node[style=node] (w) at (0.9, 0){};
				\node[style=small box] at (0, -1){$1$};
				\node[style=none](one) at (0.3, -1){};
				\node[style=node](z) at (0.9, -1){};
				\node[style=none](nil) at (0, -0.5){};
			\end{pgfonlayer}        
			\begin{pgfonlayer}{edgelayer}
				\draw(v1) to (divfst.center);
				\draw(v2) to (divsnd.center);
				\draw(divout.center) to (w);
				\draw(one.center) to (z);
			\end{pgfonlayer}
			\begin{pgfonlayer}{eqlayer}
				\draw[dashed, rounded corners](-1.2, 0.6) rectangle (-0.6, -0.6);
				\draw[dashed, rounded corners](0.6, 0.3) rectangle (1.2, -1.3);
			\end{pgfonlayer}\begin{pgfonlayer}{background}
				\draw[color=white] (-1.5, 0.2) rectangle (1.5, -0.2);
		\end{pgfonlayer}\end{tikzpicture}
	}
\]

@@

The fact that the right-hand side is a regular mono ensures that if the e-hypergraph to whom a rule is applied is acyclic,
the same property holds for the e-hypergraph obtained as the result.
%
Note however that cycles may implicitly occur due to the presence of the equivalence 
on nodes for what are called collapsing rules, i.e. where the right-hand side is a single variable. 
%
Looking again at the key example of~\cite{xxx}, consider the rule $1 \times x = x$,
which can be modelled as the DPO rule below
%
\[xxx\]
%
Intuitively, the e-term graph obtained after the application of the rule represents a recursive term such 
as $rec_x. 1 \times x$, thus highlighting the difficulty of extracting a term representation out of an e-term graph.

@@

\section*{Acknowledgments}
Hypergraphs, terms graphs, e-graphs, adhesive categories.

>>>>

Consider the term rewriting rule such $x / x \to 1$, from the key example in~\cite{WillseyNWFTP21}.
This is a problematic rule to be simulates using classical term graph rewriting, since it is not left-linear
(that is, $x$ occurs twice in the left-hand side)~\cite{AriolaKP00}.
Using e-term graphs (variables are represented as nodes that do not occur among the targets of an edge),
 it can be modelled as the DPO rule below, concisely given by the arrow $L \rightarrow R$
\[
	\hspace{.25cm}
	\xymatrix{        
		\begin{tikzpicture}[baseline=(w.base)]\begin{pgfonlayer}{nodelayer}
				\node[style=node](v1) at (-0.9, 0.3){};
				\node[style=node](v2) at (-0.9, -0.3){};
				\node[style=medium box] at (0, 0){$/$};
				\node[style=none](divfst) at (-0.3, 0.3){};
				\node[style=none](divsnd) at (-0.3, -0.3){};
				\node[style=none](divout) at (0.3, 0){};
				\node[style=node] (w) at (0.9, 0){};
			\end{pgfonlayer}        
			\begin{pgfonlayer}{edgelayer}
				\draw(v1) to (divfst.center);
				\draw(v2) to (divsnd.center);
				\draw(divout.center) to (w);
			\end{pgfonlayer}
			\begin{pgfonlayer}{eqlayer}
				\draw[dashed, rounded corners](-1.2, 0.6) rectangle (-0.6, -0.6);
			\end{pgfonlayer}\begin{pgfonlayer}{background}
				\draw[color=white] (-1.5, 0.2) rectangle (1.5, -0.2);
			\end{pgfonlayer}
		\end{tikzpicture}
		\ar@{=>}[r] &
		\begin{tikzpicture}[baseline=(nil.center)]\begin{pgfonlayer}{nodelayer}
				\node[style=node](v1) at (-0.9, 0.3){};
				\node[style=node](v2) at (-0.9, -0.3){};
				\node[style=medium box] at (0, 0){$/$};
				\node[style=none](divfst) at (-0.3, 0.3){};
				\node[style=none](divsnd) at (-0.3, -0.3){};
				\node[style=none](divout) at (0.3, 0){};
				\node[style=node] (w) at (0.9, 0){};
				\node[style=small box] at (0, -1){$1$};
				\node[style=none](one) at (0.3, -1){};
				\node[style=node](z) at (0.9, -1){};
				\node[style=none](nil) at (0, -0.5){};
			\end{pgfonlayer}        
			\begin{pgfonlayer}{edgelayer}
				\draw(v1) to (divfst.center);
				\draw(v2) to (divsnd.center);
				\draw(divout.center) to (w);
				\draw(one.center) to (z);
			\end{pgfonlayer}
			\begin{pgfonlayer}{eqlayer}
				\draw[dashed, rounded corners](-1.2, 0.6) rectangle (-0.6, -0.6);
				\draw[dashed, rounded corners](0.6, 0.3) rectangle (1.2, -1.3);
			\end{pgfonlayer}\begin{pgfonlayer}{background}
				\draw[color=white] (-1.5, 0.2) rectangle (1.5, -0.2);
		\end{pgfonlayer}\end{tikzpicture}
	}
\]

@@

The fact that the right-hand side is a regular mono ensures that if the e-hypergraph to whom a rule is applied is acyclic,
the same property holds for the e-hypergraph obtained as the result.
%
Note however that cycles may implicitly occur due to the presence of the equivalence 
on nodes for what are called collapsing rules, i.e. where the right-hand side is a single variable. 
%
Looking again at the key example of~\cite{WillseyNWFTP21}, consider the rule $1 \times x = x$,
which can be modelled as the DPO rule below
%
\[xxx\]
%
Intuitively, the e-term graph obtained after the application of the rule represents a recursive term such 
as $rec_x. 1 \times x$, thus highlighting the difficulty of extracting a term representation out of an e-term graph.

@@

%\section*{Acknowledgments}
%Hypergraphs, terms graphs, e-graphs, adhesive categories.
